Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-03-30 draft.tex
Proofread the Paper

\documentclass[journal]{style/vgtc} 			          % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data are analyzed using a priori defined hypotheses to find statistically significant correlations between variables (\emph{associations}).
%
Modern cohort studies also incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic variables and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It shows to be a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework allows for hypothesis validation and hypothesis generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image variables aims to unveil \emph{associations} by applying an iterative analysis approach.

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypothesis for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a condition of interest.
%
The data variables, that are potentially useful, are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate difference between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other variables.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques which reach beyond standard statistical methods.

@@

Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypothesis as well as trigger \emph{hypothesis generation} using data mining methods such as clustering.

@@

Hypothesis generation gains importance since the number of epidemiological variables increases and the focus shifts towards more complex relations involving more than two variables.

@@

Easy publishing of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.

@@

This data set comprises 127 variables and 2 sequences of MRI data from 2.333 patients.

@@

	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
	\item visualization techniques, which combine both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures.
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures

@@

%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
In this section we describe the epidemiological workflow and associated requirements.
% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
%
%Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the IVA supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
	\item A Hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
	\item Confounding features are identified and taken into account, for example using stratification % adjusted so that they do not influence the effect size of a attribute.
%
	\item Statistical methods such as regression analysis assess the association of selected features with the investigated condition.%are applied to measure the effect size of attributes to the outcome of interest.
	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it doesn't consider any kind of image data.
%
The extension, we suggest, are motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
%In the following we discuss the problems and challenges regarding the workflow.

@@

In longitudinal cohort design, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20 year-steps) when investigating its influence on a condition.
%
These groups depend strongly on the condition of interest and therefore there is no standard on how to categorize these.

To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature which needs to be isolated.
%
A famous example for a confounder is the correlation observable between shoe size and mortality where it can be observed that people with larger shoes have a smaller life expectation than people with small shoe size.
%
Shoe size actually is associated with gender where women have a smaller feet than men but have a longer life expectation.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.

@@

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
%MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
MRI the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort consisting of 4.308 subjects, the SHIP, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
The MRI protocol features a rich number of sequences.
%
%In addition to other MRI data, breast MRI scans were acquired for women.
%
%A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

\section{Prior and Related Work}
%Einfuehren von Helwigs Terminologie?
Designing a visualization which conveys all data aspects equally is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques need to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimension by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues is closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

@@

\paragraph{Visualizing Image and Non-Image Data. }
Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems, which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
%
They incorporated data mining methods such as dividing the data space by using a k-nearest-neighbor technique and the PCA.
%
Steenwijk and colleagues employ a relational database to organize the data to visualize subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every subject group divided by parameter as individual cohort.

\paragraph{Visualizing Heterogenous Non-Image Data.}
Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
%
They are useful to gain an overview over numerous variables and their distributions.
%This technique is also useful to gain an overview over numerous variables and their distributions.
%
Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
%
%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
Their findings yielded a \emph{Concept Map} which linked cancer-related associations via graph edges.
%From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
Distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Derived from this we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related difference \cite{Hermann2014}. 
%
The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%

>>>>

\documentclass[journal]{style/vgtc} 			          % final (journal style)
%\documentclass[review,journal]{style/vgtc}         % review (journal style)
%\documentclass[widereview]{style/vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{style/vgtc}       % preprint (journal style)
%\documentclass[electronic,journal]{style/vgtc}     % electronic version, journal
%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier. Further, ``electronic'' includes
%% hyperreferences for more convenient online viewing.
% TODO: Incomplete data: Cramers V kann nur für bestimmte mindestgrößen funktionieren - entweder Daten unvollständig oderz u wneige Probanden in einer Gruppe
%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors.
%
The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
Modern cohort studies incorporate medical image data.
%
%Analyzing these data requires image segmentation, extraction of key figures, and shape-based subject grouping.
%
The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach, that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%incorporating coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%Multiple coordinated views from information visualization and 3D-shape renderings are combined with epidemiological data representations such as Pivot Tables.
%
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and -generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond, are based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
On the other hand, observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examinations results, which may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of persons randomly selected to avoid any bias.
%
Due to this random selection, the majority of the cohort does not suffer from a specific health problem.
%
A large size of the cohort is therefore essential to investigate difference between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
%These data needs to be segmented to analyze its features for correlations with non-image data.
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%The segmentation of these data would allow to analyze shapes of anatomical structures and to correlate them with other features.
%These data needs to be concurrently analyzed with the non-image data.
%
%This requires their segmentation and the derivation of features such as kidney volume, curvature of the spine or tissue texture of the liver.
%
%Manual segmentation is not feasible since it leads to a high variability which is at odds with the high reproducibility required in epidemiology research.
%
Semi-automatic techniques are more promising but also challenging since the employed modalities, such as Magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques, which reach beyond standard statistical methods.

@@

Intended as an extension to the well established epidemiological tools it provides a way to rapidly validate hypotheses as well as trigger \emph{hypothesis generation} using data mining methods, such as clustering.

@@

\emph{Hypothesis generation} gains importance since the number of epidemiological features increases and the focus shifts towards more complex relations involving more than two features.

@@

Easy exchange of developed methods driven by modern web technologies intends to trigger a fast feedback loop between us and the epidemiologists.

@@

This data set comprises 127 features and 2 sequences of MRI data from 2.333 patients.

@@

	\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow,
	%\item an Interactive Visual Analysis workflow for analyzing image-based epidemiological data including both hypothesis-driven and for hypothesis-generation based on a characterization of the standard epidemiological workflow
	\item visualization techniques, which incorporate both information visualization and 3D rendering of organ shapes as well as combining them with epidemiological graphics and key figures,
	\item highlighting interesting subject groups and feature associations using shape-based clustering and statistical contingency measures,

@@

%In this section we give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges. %study data to identify the problems we address in this paper.
In this section we describe the epidemiological workflow and associated requirements.
% give insight into the epidemiological workflow when analyzing cohort study data and derive several challenges
%
%Epidemiological Outcome - 5Ds:
% Death
% Disease
% Discomfort
% Disability
% Dissatisfaction
%
\subsection{Epidemiological Workflow} \label{EpidemiologicalWorkflow}
\begin{figure}[htb]
 \centering
 \includegraphics[width=3.0in]{figures/workflow_comparison}
 \caption{\emph{IVA} tools are able to complement parts of the epidemiological workflow, not replace it. The appropriate combination of statistical- and interactive driven analysis shows promising potential to unveil the information in the data. (a) shows the standard epidemiological workflow, (b) the \emph{IVA} supported one. The iterative red highlighted part is called the \emph{IVA Loop} and is described in more detail in Figure~\ref{fig:InteractionLoop}.}
  \label{fig:WorkflowComparison}
\end{figure}
%Epidemiology diversity is reflected in the different experts working at cohort studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
The diversity of epidemiology is reflected in the different experts who work at cohort studies, ranging from specialized doctors, to medical computer scientists with focus on biometrics, and statisticians.
%Many different experts work at epidemiological studies, ranging from specialized doctors to medical computer scientists with focus on biometrics and statisticians.
%
Epidemiologists follow a workflow mainly driven by statistic tools to validate hypotheses about disease-specific risk factors.
%
Following Thew and colleagues, the workflow can be characterized as follows \cite{Thew2009}:
%
\begin{enumerate}
%	\item Hypotheses are based most commonly on observations made by clinicians in their daily routine.
	\item A hypothesis is derived from observations made by clinicians in their daily routine.
%
	\item A set of features depicting conditions affected by the hypothesis is compiled accordingly.
	%\item Confounding features are adjusted so that they do not influence the effect size of an attribute.
	\item Confounding features are identified and taken into account (for example using stratification). % adjusted so that they do not influence the effect size of a attribute.
%
	\item Statistical methods, such as regression analysis, assess the association of selected features with the investigated disease.%are applied to measure the effect size of attributes to the outcome of interest.
	%\item Statistical methods such as regression analysis are applied to measure the effect size of attributes to the outcome of interest.
\end{enumerate}
The workflow is shown in Figure~\ref{fig:WorkflowComparison} (a) and serves as orientation for our approach although it doesn't consider any kind of image data.
%
The extension we suggest, is motivated by the potential of image data and the attempt to support \emph{hypothesis generation}.
%In the following we discuss the problems and challenges regarding the workflow.

@@

In longitudinal cohort design, grouping subjects using epidemiological features is essential in order to allow per-group risk determination.
%
%Grouping is carried out hypothesis driven.
Grouping depends on the underlying hypothesis.
%
Age for example is divided into groups (e.g. in 20-year-steps) when investigating its influence on a condition.
%
These groups strongly depend on the condition of interest and therefore there is no standard on how to categorize them.

To determine, whether a subject is prone to be affected by a certain disease, \emph{relative risks} are expressed through the evaluation of p-values, which indicate statistical significance.
%
Statistical correlations are prone to \emph{confounding}, meaning that the association of two features are influenced by a third feature, which needs to be isolated.
%
A famous example for a confounder is the correlation observable between shoe size and mortality where it can be observed that people with larger shoes have a smaller life expectation than people with small shoe size.
%
Shoe size actually is associated with gender where women have a smaller feet than men but have a longer life expectation.
% Statistical correlations are prone to \emph{confounding}, meaning that two features are dependent and therefore should be normalized with respect to each other.
%
%Statistics tools such as \texttt{SPSS}\footnote{Product of IBM; \url{www.ibm.com/software/de/analytics/spss/}} and \texttt{Stata}\footnote{Product of Stata Corp.; \url{http://www.stata.com/}} play a major role for analyzing epidemiological data.

@@

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
%MRI data is more expensive to obtain than CT data but does not affect the subjects health and is therefore the main method for collecting cohort study imaging data.
Therefore is MRI the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is kept, if possible, on the initial software and hardware version during a longitudinal study to ensure comparability in and between acquisition cycles.

@@

\subsection{The Study of Health in Pomerania (SHIP)}
After the pioneering Rotterdam study (started in 1990) several MR imaging studies initiatives have evolved.
%
They slightly differ in clinical focus, acquired data and epidemiological research questions.
%
Starting in 1997 with a cohort consisting of 4.308 subjects, the \texttt{SHIP}, located in northern Germany, aims to characterize health and disease in the widest range possible \cite{Volzke2011}.
%
Data is collected without focus on a group of diseases.
%
This allows the data set to be queried regarding many different diseases and conditions.
%
Subjects were examined in a 5-year time span, continuously adding new parameters including MRI scans in the last iteration \cite{Hegenscheid2009}.
%
The MRI protocol features a rich number of sequences.
%
%In addition to other MRI data, breast MRI scans were acquired for women.
%
%A second cohort \texttt{SHIP-Trend} was established in 2008 to acquire data about a younger population.
A second cohort \texttt{SHIP-Trend} was established in 2008.
%
The protocols for examining the subjects between \texttt{SHIP} and \texttt{SHIP-Trend} remained the same, making them comparable.
%
The overall examination time for each person attending the study is two days.

\section{Prior and Related Work}
%Einfuehren von Helwigs Terminologie?
Designing a visualization, which conveys all data aspects equally, is challenging.
%
Given the number of features of epidemiological data sets and their different manifestations, the strength of different visualization techniques needs to be combined \cite{Buja91, Konyha2009}.
%
The principal component analysis (PCA) and similar techniques are able to reduce the dimensions by extracting most expressive components, but make the influence of each variable hard to convey.
\\\\
Their focus on hypothesis generation based on parallel assessment of multiple data features makes the work of Turkay and colleagues closest to ours albeit our emphasis on processing medical image data and variables with categorical manifestations \cite{Turkay2013}.
%
Their methods aim to amplify a hypothesis generation process for analyzing data of a Norwegian aging study.
%
%Statistical measures of continuous variables such as mean, standard deviation, skewness, or inter-quartile range are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures.
Statistical measures of continuous variables, such as mean, standard deviation, skewness, or inter-quartile range, are used to create \emph{dimension plots} that make them comparable with respect to the derived descriptive measures, such as voxel number of a segmented structure.
%
%These transform dimensions into data points and make them comparable with respect to the derived descriptive measures, making them comparable.
%
%This not only allows for comparing all continuous variables in a single plot but make their distribution change comprehensible.
%
%This requires a good descriptive measure which captures the kind of change the user is interested in or which reflects unexpected data behavior.
%
%The technique was applied to features obtained by segmenting the brain into 45 parts and measure the voxel number, volume and properties of the intensity values.
%
The method is strongly dependent on the descriptive measures of the epidemiological factors.

@@

\paragraph{Visualizing Image and Non-Image Data. }
Gresh and colleagues proposed \texttt{WEAVE}, one of the first systems, which concurrently analyzed medical image and non-image data using linked views \cite{Gresh2000}.
%
Blaas and colleagues presented a similar system which analyzed medical image data and variables derived from them using views from the feature- and physical space \cite{Blaas2007}.
%
They incorporated data mining methods such as dividing the data space by using a k-nearest-neighbor technique and the PCA.
%
Steenwijk and colleagues employ a relational database to organize the data for visualizing subject data using linked views such as parallel coordinates, scatterplots and time plots \cite{Steenwijk2010}.
%
Zhang and colleagues provide a web-based system for analyzing subject groups with linked views and batch-processing capabilities for categorizing new subject entries into the data set \cite{Zhang2012}.
%
Their understanding of a cohort differs from the understanding of the term in an epidemiological context by denoting every parameter divided subject group as individual cohort.

\paragraph{Visualizing Heterogenous Non-Image Data.}
Generalized Pairs Plots (\texttt{GPLOM´S}) are an information visualization technique comparing heterogenous variables pairwise using a plot-matrix grouped by type \cite{Francois2013}.
%
They are useful to gain an overview over numerous variables and their distributions.
%This technique is also useful to gain an overview over numerous variables and their distributions.
%
Histograms, bar charts, scatterplots and heat maps are used to visualize variable combinations with regard to their type.
%
%Brushing, linking and filtering are feasible with \texttt{GPLOMS}, but have limitations such as making only one category brushable at a time.
%
%We applied this technique to our data and it shows promising potential for simultaneously visualizing many different variables. %but does not fit in the scope of this paper.
%
%The inspiration on the chosen visualization techniques stems from this publication.
%
Dai and colleagues explored risk factors by incorporating choropleth maps of epidemiological features (e.g., mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%A similar approach was taken by Dai and colleagues for risk factor exploration as they also incorporate choropleth maps of epidemiological factors (e.g. mortality rates in a region) with parallel coordinates, bar charts and scatterplots with integrated regression lines \cite{Dai2005}.
%
Their findings yielded a \emph{Concept Map}, which linked cancer-related associations via graph edges.
%From their findings regarding theassociationn of cancer-related socio-demographic factors are drawn in a \emph{Concept Map} where related factors are connected via graph-edges.
Chui and colleagues visualized associations in time-dependent epidemiological data using time-series plots highlighting risk factors differences in age and gender \cite{Chui2011}.

\paragraph{Commercial Data Visualization.}
Commercial systems such as \texttt{Tableau}\footnote{Owned by Tableau Software; \url{www.tableausoftware.com}} or \texttt{Spotfire}\footnote{Owned by TIBCO; \url{spotfire.tibco.com}} provide a rich user interface that enable Visual Analytics approach without the need of writing any code.
%
With little effort, linked views can be created, but the data processing possibilities such as derivation of new variables or the 3D rendering capabilities are very limited.
%
These systems share limitations in extensibility to a specific problem domain.
%Spotfire offers the possibilities to interact with the \texttt{R} statistical programming language 
% Image Data
% \paragraph{Visualizing Image Data.}
\paragraph{Visualizing Shape Variance.}
Comparing tissue between many subjects in an epidemiological context requires methods which allow for shape variance visualizations.
%
Caban and colleagues investigated the suitability of variance visualizations of shape distribution models and concluded in their user study that users favor spherical glyph representations over deformation grids and likelihood volumes \cite{Caban2011}.
%
The distribution of shapes in a space derived from a PCA is plotted by Busking and colleagues in a 2D-projected plane of the space \cite{Busking2010a}.
%
Interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Interpolation is carried out by mesh morphing.
%
Distance to the mean shape is color-coded.
%Differences between structures are highlighted using color mapping of the difference to the mean shape, but is rather hard to recognize due to small renderings of each subject in the shape-space.
%Via mesh morphing interpolated views can be created by the user in a separate view as well as comparisons in a contour view.
%
Derived from this, we incorporate the idea of combining 3D-shape rendering with information visualization techniques.
%
%Applying our data sets to this technique yielded a cluttered shape space due to the many subjects.
%
%The data needs to abstracted or summarized in order to work in this context.
%
Hermann and colleagues identify local deformation changes by investigating shape-related difference \cite{Hermann2014}. 
%
The user specifies a deformation of interest and showing corresponding changes in the shape using covariance tensors.
%
This method allowed for rapid hypotheses validation and was able to reproduce textbook knowledge.
%
%By plotting p-values in ventricle surfaces, Chou and colleagues were able to map disease-associated values directly on a 3D tissue representations \cite{Chou2009}.
%
%This requires a geographic colocation of associated features.
%
